---
title: "Ken Griffey, Jr. Career Trajectory"
author: "Martin Frigaard"
output: 
  html_document: 
    toc: yes
    toc_depth: 6
    highlight: zenburn
    theme: united
    df_print: paged
---


```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(tidyverse)
# figures folder
fs::dir_create("figs/")
## Global options
options(max.print = "75")
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  comment = "#> ",
  message = FALSE,
  warning = FALSE,
  fig.path = "figs/",
  fig.retina = 2,
  fig.width = 9, 
  fig.height = 7
)
knitr::opts_knit$set(width = 75)
```

```{r ipsum, message=FALSE, warning=FALSE, cache=FALSE, echo=FALSE}
library(hrbrthemes)
library(tidyverse)
update_geom_font_defaults(font_rc)
```

# Motivation

This repo contains an example career trajectory calculation for Ken Griffey, Jr. I've borrowed heavily from [Analyzing Baseball Data with R, 2nd Edition](https://www.amazon.com/Analyzing-Baseball-Data-Second-Chapman/dp/0815353510) and [Frontiers in Major League Baseball](https://www.springer.com/gp/book/9781441908308).

## Packages 

The two primary packages we will use are `tidyverse` and `Lahman`

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(Lahman)
```


## Get player statistics 

```{r kgjr_id}
kgjr_id <- Lahman::People %>% 
  filter(nameFirst == "Ken" & nameLast == "Griffey" & birthYear == 1969) %>% 
  pull(playerID)
kgjr_id
```

We need to change the value of missing statistics from the `Batting` table with 0s, which we can do below. 

`SF` = Sacrifice flies

`HBP` = Hit by pitch

```{r replace-missing-Batting}
Batting <- Lahman::Batting %>% 
  tidyr::replace_na(list(SF = 0, HBP = 0))
```

Now we write a custom function for getting the player stats with a given `playerID`. 

```{r get_stats}
get_stats <- function(player.id) {
  Batting %>%
    dplyr::filter(playerID == player.id) %>%
    dplyr::inner_join(Master, by = "playerID") %>%
    dplyr::mutate(birthyear = ifelse(birthMonth >= 7, birthYear + 1, birthYear), 
      Age = yearID - birthyear, 
      SLG = (H - X2B - X3B - HR + 2 * X2B + 3 * X3B + 4 * HR) / AB,
      OBP = (H + BB + HBP) / (AB + BB + HBP + SF),
      OPS = SLG + OBP) %>%
dplyr::select(Age, SLG, OBP, OPS)
}
```

Now we can view Ken Griffey Jr.'s (`KGJr`) stats over his age (in years).

```{r KGriffeyJr}
KGriffeyJr <- get_stats(kgjr_id)
KGriffeyJr %>% rmarkdown::paged_table()
```

Now we'll take a look at `KGJr`'s OPS vs. his age.

## Age vs. slugging percentage and on-base percentage

The OPS is the sum of a player's slugging percentage and on-base percentage. When we plot it against `KGJr`s age, we can see there is an increase until about age 24, then a gradual decline until age 40.

```{r example-fig}
KGriffeyJr %>% 
  ggplot2::ggplot(aes(x = Age, y = OPS)) + 
  ggplot2::geom_point() + 
  hrbrthemes::theme_ipsum_tw(axis_title_size = 13, axis_text_size = 12) + 
  ggplot2::labs(title = "Ken Griffey Jr Career Trajectory", 
                subtitle = "Age vs. slugging percentage + on-base percentage")
```

## Modeling career performance 

We will borrow the model from [Analyzing Baseball Data with R, 2nd Edition](https://www.amazon.com/Analyzing-Baseball-Data-Second-Chapman/dp/0815353510) and fit a quadratic curve to `OPS` and `Age`. 

> `A + B(Age − 30) + C(Age − 30)^2`

Where `A`, `B`, and `C` are the following, 

`A` = the predicted value of OPS when the player is 30 years old  

`30 - (B/2C)` = The age where the player is estimated to have his peak batting performance during his career (PEAK AGE).

`A - (B^2/4C)` = maximum value of the curve or the estimated largest `OPS` of the player over his career.

`C` = the degree of curvature in the quadratic function (the larger the number, the steeper the climb and subsequent decline)

Written as a formula for R to interpret, it looks like this:

`OPS ~ I(Age - 30) + I((Age - 30)^2)`

```{r fit}
fit <- lm(OPS ~ I(Age - 30) + I((Age - 30)^2), data = KGriffeyJr)
fit
```

Now we get the coefficients for `fit`

```{r b}
b <- coef(fit)
b
```

Next we need the age_max, or the `30 - (B/2C)`. We can get this by using the 2nd and 3rd element stored in `b` (which are the coefficients for `I(Age - 30)` and `I((Age - 30)^2)`).

age max = `30 - [I(Age - 30)] / [I((Age - 30)^2)] / 2`

```{r age_max}
b[2]
b[3]
age_max <- 30 - b[2] / b[3] / 2
age_max
```

With our `age_max`, now we just need the `max_value` or `A - (B^2/4C)` which is `b[1] - b[2] ^ 2 / b[3] / 4` or 

max value = `[(Intercept)] - [I(Age - 30)]^2 / [I((Age - 30)^2)] / 4`

```{r max_value}
b[1]
max_value <- b[1] - b[2] ^ 2 / b[3] / 4
max_value
```

Finally we combine them into a list. 

```{r list-fit-age_max-max_value}
list(fit = fit, age_max = age_max, max_value = max_value)
```


```{r fit_model}
fit_model <- function(d) {
  
  fit <- lm(OPS ~ I(Age - 30) + I((Age - 30)^2), data = d)
  
  b <- coef(fit)
  
  age_max <- 30 - b[2] / b[3] / 2
  
  max_value <- b[1] - b[2] ^ 2 / b[3] / 4
  
  list(fit = fit, age_max = age_max, max_value = max_value)
}
```

```{r KGJr_fit_model}
KGJr_fit_model <- fit_model(KGriffeyJr)
coef(KGJr_fit_model$fit)
```

```{r KGJr_fit_model-age_max-max_value}
c(KGJr_fit_model$age_max, KGJr_fit_model$max_value)
```

Best fitting curve is,

`0.97276346 − 0.01198292(Age − 30) − 0.00238516(Age − 30)^2`

According to this model: 

- `KGJr` peaked at age `27.49`  

- His maximum OPS for the curve is stored in `KGJr_fit_model$max_value` and is estimated to be 0.99. 

- The estimated curvature parameter is the `I((Age - 30)^2)` value: `-0.00238516`.

- `KGJr` decrease in `OPS` between his peak age (``27.49``) and one year older is `-0.00238516`.

We can plot this below by creating a few labels for the `Peak Age` and `Max OPS`, and adding 

```{r plot-curve}
peak_age_label_y <- mean(KGriffeyJr$OPS)
# peak_age_label_y
peak_age_label_x <- KGJr_fit_model$age_max + 1.5
# peak_age_label_x
max_value_label_y <- KGJr_fit_model$max_value + 0.05
# max_value_label_y
max_value_label_x <- min(KGriffeyJr$Age) + 1
# max_value_label_x
KGriffeyJr %>% 
  ggplot2::ggplot(aes(x = Age, y = OPS)) + 
  ggplot2::geom_point() + 
  ggplot2::geom_smooth(method = "lm", se = FALSE, size = 1.5, color = "salmon", 
              formula = y ~ poly(x = x, degree = 2, raw = TRUE)) +
  ggplot2::geom_vline(xintercept = KGJr_fit_model$age_max, linetype = "dashed", 
             color = "royalblue") +
  ggplot2::geom_hline(yintercept = KGJr_fit_model$max_value, linetype = "dashed", 
             color = "darkgreen") + 
  ggplot2::annotate(geom = "text", x = c(peak_age_label_x), 
                          y = c(peak_age_label_y), 
           label = c("Peak age"), size = 5, color = "royalblue") + 
  ggplot2::annotate(geom = "text", x = c(max_value_label_x), 
                          y = c(max_value_label_y), 
           label = c("Max OPS"), size = 5, color = "darkgreen") +
  hrbrthemes::theme_ipsum_tw(axis_title_size = 13, axis_text_size = 12) +
  ggplot2::labs(title = "Ken Griffey Jr Career Trajectory", 
                subtitle = "Age vs. slugging percentage + on-base percentage")
```






## Deploy

Run this code to knit an html file and move it to `docs` folder. 

```{r deploy-to-github, results='hide', message=FALSE, warning=FALSE}
# create docs folder
fs::dir_create("docs")
# knit and move to docs folder
knitr::knit(input = "index.Rmd", output = "docs/index.html")
```

